Using HyperNEAT vs. Parameter Optimization to Create Quadruped Robot Gaits

### 2011-04-13 paper_ecal/fitnessEvaluation.tex
More edits from reviews, made figures larger

\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{.7}{A Nintendo Wii
  remote provided the location of the robot by tracking the infrared
  LED mounted on the robot's antenna.  The LED position was measured
  in pixels and transmitted from the Wii remote to the robot via
  bluetooth.}

@@

As mentioned earlier, the metric for evaluating gaits was their
average speed, which was measured as the Euclidian distance the robot
moved during a 12-second run. For the manual and parameterized gaits, the fitness was this value. The HyperNEAT gaits stressed the motors more than the other gaits, so to encourage gaits that did not tax the motors we penalized gaits that caused the servos to stop responding. The penalty was to set the fitness to half of the distance the robot actually traveled. We tested whether the servos were responding after each gait by commanding them to specific positions and checking whether they actually moved to those positions. This test had the additional benefit of rewarding those gaits that did not flip the robot into a position where it could not move its legs, which HyperNEAT also did more than the other learning methods. Because the fitness of HyperNEAT gaits were often halved, in results we compare actual distance traveled in addition to fitness for the best gaits produced by each class of gait-generating algorithms.

>>>>

\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
  remote provided the location of the robot by tracking the infrared
  LED mounted on the robot's antenna.  The LED position was measured
  in pixels and transmitted from the Wii remote to the robot via
  bluetooth.}

@@

As mentioned earlier, the metric for evaluating gaits was their
average speed, which was measured as the Euclidian distance the robot
moved during a 12-second run. For the manual and parameterized gaits,
the fitness was this value. The HyperNEAT gaits stressed the motors
more than the other gaits, so to encourage gaits that did not tax the
motors we penalized gaits that caused the servos to stop
responding. When the servos stopped responding they could, in nearly
all cases, be restarted by cycling power, though over the course of
this study we did have to replace four servos that were damaged.
The penalty was to set the fitness to half of the distance
the robot actually traveled. We tested whether the servos were
responding after each gait by commanding them to specific positions
and checking whether they actually moved to those positions. This test
had the additional benefit of rewarding those gaits that did not flip
the robot into a position where it could not move its legs, which
HyperNEAT also did more than the other learning methods. Because the
fitness of HyperNEAT gaits were often halved, in results we compare
actual distance traveled in addition to fitness for the best gaits
produced by each class of gait-generating algorithms.

### 2011-04-13 paper_ecal/introduction.tex
More edits, 75% done :)
More edits from reviews, made figures larger

\acmFig{robot_whitebg.jpg}{.85}{The quadruped robot. The translucent
  parts were created with 3D-printing technology. Videos of the gaits
  evolved for this robot can be viewed at www.anonymous.url.}

@@

A final motivation of this problem is to simply evolve effective gaits
for a physical robot. Because HyperNEAT performed well in simulation,
it is interesting to test whether it can produce a fast gait for a
physical robot. It is additionally interesting to test how more
traditional gait optimization techniques compete with evolutionary
algorithms when evolving in hardware.

>>>>

\acmFig{robot_whitebg.jpg}{.75}{The quadruped robot. The translucent
  parts were created with 3D-printing technology. Videos of the gaits
  evolved for this robot can be viewed at www.anonymous.url.}

@@

A final motivation of this problem is to simply evolve effective gaits
for a physical robot. Because HyperNEAT has performed well in
simulation, it is interesting to test whether it can produce a fast
gait for a physical robot. It is additionally interesting to test how
more traditional gait optimization techniques compete with
evolutionary algorithms when evolving in hardware.

### 2011-04-13 paper_ecal/relatedWork.tex
More edits from reviews, made figures larger

Various machine learning techniques have proved to be effective at
generating gaits for legged robots. Kohl and Stone presented a policy
gradient reinforcement learning approach for generating a fast walk on
legged robots\citep{kohl}, which we implemented for comparison. Others
have evolved gaits for legged robots, producing competitive
results~\citep{chernova2005evolutionary, hornby2005autonomous, zykov,
  clune2009evolving, clune2011performance, clune2009hybrid,
  clune2009sensitivity, tellez2006evolving, valsalam2008modular}. In
fact, an evolved gait was used in the first commercially-available
version of Sony's AIBO robot~\citep{hornby2005autonomous}. Except for
work with HyperNEAT~\citep{clune2009evolving, clune2011performance,
  clune2009hybrid, clune2009sensitivity}, the previous evolutionary
approaches have helped evolution exploit the regularity of the problem
by manually decomposing the task. Experimenters have
to choose which legs should be coordinated, or otherwise facilitate
the coordination of motion. Part of the motivation of this paper
is to see if the space of regularities can be explored automatically,
which has previously performed well~\citep{valsalam2008modular}, and to
perform a direct comparison on the same robot to HyperNEAT.

>>>>

Various machine learning techniques have proved to be effective at
generating gaits for legged robots. Kohl and Stone presented a policy
gradient reinforcement learning approach for generating a fast walk on
legged robots \citep{kohl}, which we implemented for comparison. Others
have evolved gaits for legged robots, producing competitive
results \citep{chernova2005evolutionary, hornby2005autonomous, zykov,
  clune2009evolving, clune2011performance, clune2009hybrid,
  clune2009sensitivity, tellez2006evolving, valsalam2008modular}. In
fact, an evolved gait was used in the first commercially-available
version of Sony's AIBO robot~\citep{hornby2005autonomous}. Except for
work with HyperNEAT \citep{clune2009evolving, clune2011performance,
  clune2009hybrid, clune2009sensitivity}, the previous evolutionary
approaches have helped evolution exploit the regularity of the problem
by manually decomposing the task. Experimenters have
to choose which legs should be coordinated, or otherwise facilitate
the coordination of motion. Part of the motivation of this paper
is to see if the space of regularities can be explored automatically,
which has previously performed well \citep{valsalam2008modular}, and to
perform a direct comparison on the same robot to HyperNEAT.

### 2011-04-13 paper_ecal/results.tex
More edits from reviews, made figures larger

The results for the parameterized gaits are shown in
\figref{std_error} and \tabref{results}.  The best overall gait for
the parameterized methods was found by linear regression, which also
had the highest average performance. The Nelder-Mead simplex also
performed quite well on average.  The other local search methods did not
outperform random search; however, all methods did manage to explore
enough of the parameter space to significantly improve on the previous
hand-coded gait in at least one of the three runs.  No single strategy
consistently beat the others: for the first trial Linear Regression
produced the fastest gait at 27.58 body lengths/minute, for the second
a random gait actually won with 17.26, and for the
third trial the Nelder-Mead simplex method attained the fastest gait
with 14.83.

@@

\acmFig{std_error}{1}{Average results ($\pm $ SE) for each of the
  parameterized learning methods.  Linear regression found the fastest
  overall gait and had the highest average, followed by Nelder-Mead
  simplex. Other methods did not outperform a random strategy.}

@@

The results for the gaits evolved by HyperNEAT are shown in
\figref{hnResults} and \tabref{results}.  Overall the HyperNEAT gaits
were the fastest by far, beating all the parameterized models when
comparing either average or best gaits.  We believe that this is
because HyperNEAT was allowed to explore a much richer space of
motions, but did so while still utilizing symmetries when
advantageous.  The single best gait found during this study had a
speed of 45.72 body lengths / minute, 66\% better than the best
non-HyperNEAT gait and 8.9 times faster than the hand-coded gait.
\figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
HyperNEAT gait that had high fitness.  The pattern of motion is both
complex (containing multiple frequencies and repeating patterns across
time) and regular, in that patterns of multiple motors are
coordinated.

The evaluation of the gaits produced by HyperNEAT was more noisy than for the parameterized gaits, which made learning difficult. For example, we tested an example HyperNEAT generation-champion gait 11 times and found that its mean performance was 26 body lengths per minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many effective HyperNEAT gaits were not preserved across generations because if performance in one trial was poor, the genome was unlikely to be selected for. We believe the slope of the HyperNEAT learning curve would be steeper if the noise in the evaluations could be reduced.

\acmFig{hnResults}{.8}{Average fitness ($\pm $ SE) of the highest
  performing individual in the population for each generation of
  HyperNEAT runs. The fitness of many high-performing HyperNEAT gaits
  were halved if the gait overly stressed the motors (see text),
  meaning that HyperNEAT's true performance without this penalty would
  be much higher.}

@@

\acmFig{neat_110115_211410_00000_002_filt_zoom}{.8}{Example of one
  high-performance gait produced by HyperNEAT showing
  commands for each of nine motors.  Note the complexity of the motion
  pattern. Such patterns were not possible with the parameterized
  SineModel5, nor would they likely be contrived by a human designing
  a different low-dimensional parameterized motion model.}

>>>>

The results for the parameterized gaits are shown in
\figref{std_error} and \tabref{results}.  A total of 1217 hardware
fitness evaluations were performed during the learning of
parameterized gaits, with the following distribution by learning
method: 200 random, 234 uniform, 284 Gaussian, 174 gradient, 172
simplex, 153 linear regression.  The number of runs varies because
each run plateaued at its own pace. The best overall gait for
the parameterized methods was found by linear regression, which also
had the highest average performance. The Nelder-Mead simplex also
performed quite well on average.  The other local search methods did not
outperform random search; however, all methods did manage to explore
enough of the parameter space to significantly improve on the previous
hand-coded gait in at least one of the three runs.  No single strategy
consistently beat the others: for the first trial Linear Regression
produced the fastest gait at 27.58 body lengths/minute, for the second
a random gait actually won with 17.26, and for the
third trial the Nelder-Mead simplex method attained the fastest gait
with 14.83.

@@

\acmFig{std_error}{1}{Average results ($\pm $ SE) for each of the
  parameterized learning methods, computed over three separately
  initialized runs.  Linear regression found the fastest overall gait
  and had the highest average, followed by Nelder-Mead simplex. Other
  methods did not outperform a random strategy.}

@@

The results for the gaits evolved by HyperNEAT are shown in
\figref{hnResults} and \tabref{results}.  A total of 540 evaluations
were performed for HyperNEAT, 180 in each of three runs.
Overall the HyperNEAT gaits
were the fastest by far, beating all the parameterized models when
comparing either average or best gaits.  We believe that this is
because HyperNEAT was allowed to explore a much richer space of
motions, but did so while still utilizing symmetries when
advantageous.  The single best gait found during this study had a
speed of 45.72 body lengths / minute, 66\% better than the best
non-HyperNEAT gait and 8.9 times faster than the hand-coded gait.
\figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
HyperNEAT gait that had high fitness.  The pattern of motion is both
complex (containing multiple frequencies and repeating patterns across
time) and regular, in that patterns of multiple motors are
coordinated.

The evaluation of the gaits produced by HyperNEAT was more noisy than
for the parameterized gaits, which made learning difficult. For
example, we tested an example HyperNEAT generation-champion gait 11
times and found that its mean performance was 26 body lengths per
minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many
effective HyperNEAT gaits were not preserved across generations
because if performance in one trial was poor, the genome was unlikely
to be selected for. The HyperNEAT learning curve would be smoother if
the noise in the evaluations could be reduced or more than one
evaluation per individual could be afforded.

\acmFig{hnResults}{1}{Average fitness ($\pm $ SE) of the highest
  performing individual in the population for each generation of
  HyperNEAT runs. The fitness of many high-performing HyperNEAT gaits
  were halved if the gait overly stressed the motors (see text),
  meaning that HyperNEAT's true performance without this penalty would
  be much higher.  Each generation involved nine evaluations, one for
  each of the nine individuals, so the total evaluations per HyperNEAT
  run was 180.}

@@

\acmFig{neat_110115_211410_00000_002_filt_zoom}{1}{Example of one
  high-performance gait produced by HyperNEAT showing
  commands for each of nine motors.  Note the complexity of the motion
  pattern. Such patterns were not possible with the parameterized
  SineModel5, nor would they likely be contrived by a human designing
  a different low-dimensional parameterized motion model.}
